---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tinycensus

<!-- badges: start -->
<!-- badges: end -->

`tinycensus` is a lightweight, metadata-driven interface to the [US Census Bureau API](https://www.census.gov/data/developers.html).
It is designed to stay small while still covering a broad set of aggregate datasets, with first-class support for ACS, decennial census products, PEP, CBP, migration flows, time-series endpoints, and related discovery metadata.

The package is built around product-specific interfaces rather than one generic user-facing query function.
The goal is to make common Census workflows easy without pulling in a large dependency stack.

The basic pattern is:

1. discover a dataset, table, or variable with the metadata helpers
2. retrieve data with a product-specific wrapper
3. optionally join geometry with `tinytiger`

## Installation

You can install the development version of tinycensus from GitHub with:

``` r
# install.packages("pak")
pak::pak("christopherkenny/tinycensus")
```

## What it does

- Discovers Census datasets from the live API catalog
- Retrieves metadata for variables, tables, geographies, and example queries
- Provides product-specific retrieval for ACS, decennial census, PEP, CBP, flows, and time-series datasets
- Searches variables and retrieves encoded value metadata
- Accepts flexible geography inputs like `"NY"`, `"New York"`, and `"36"`
- Returns optional `sf` output through `tinytiger`

## Core workflows

The main entry points are:

- `tc_get_acs()` for ACS products
- `tc_get_decennial()` for decennial census products
- `tc_get_pep()` for population estimates
- `tc_get_cbp()` for County Business Patterns
- `tc_get_flows()` for ACS migration flows
- `tc_get_timeseries()` for discovery-catalog time-series endpoints

For metadata discovery, start with `tc_datasets()`, `tc_variables()`, `tc_tables()`, and `tc_search_variables()`.

## API key

You can use many endpoints without a key for light exploration, but a Census API key is recommended for regular usage.

``` r
tinycensus::tc_set_key("YOUR-KEY")
```

To write the key to a `.Renviron` file for future sessions, set `install = TRUE`
and supply the target path explicitly:

``` r
tinycensus::tc_set_key(
  "YOUR-KEY",
  install = TRUE,
  r_env = file.path(Sys.getenv("HOME"), ".Renviron")
)
```

You can also check whether a key is available:

```{r key}
library(tinycensus)

tc_has_key()
```

## Discover datasets

```{r datasets}
tc_datasets(year = 2024, family = "acs")[1:5, c("year", "dataset", "title")]
```

## Flexible geography inputs

State inputs can be abbreviations, names, or FIPS codes directly in the query interface:

```{r normalize}
tc_get_acs(
  year = 2024,
  variables = "B01001_001E",
  geography = "state",
  state = c("NY", "Delaware", "36")
)
```

County normalization is geography-vintage aware.
That means `tinycensus` uses the county definitions implied by the dataset rather than assuming that the request year alone determines valid county codes.
For example, ACS 2024 uses current county equivalents, while decennial 2020 products keep 2020 county definitions.
If you need to override that behavior for a specific dataset, use `geography_vintage =`.

The same interface also works with less-common geographies when the dataset supports them, such as places, congressional districts, metropolitan areas, and school districts.

## Retrieve Census data

The ACS wrapper makes common requests compact:

```{r acs}
tc_get_acs(
  year = 2024,
  variables = "B19013_001E",
  geography = "state",
  state = c("NY", "Delaware")
)
```

You can do the same with a decennial dataset:

```{r decennial}
tc_get_decennial(
  year = 2020,
  dataset = "pl",
  variables = "P1_001N",
  geography = "county",
  state = "Delaware"
)
```

You can also request a full ACS table directly:

```{r acs-table}
tc_get_acs(
  year = 2024,
  table = "B01001",
  geography = "state",
  state = "Delaware"
)
```

## Inspect metadata

Metadata helpers make it easier to explore unfamiliar datasets before you query them.
For most workflows, `tc_tables()` is the best way to browse table-level metadata; `tc_groups()` remains available as a lower-level helper for the raw Census API concept.

```{r metadata}
vars <- tc_variables("acs/acs5", 2024)
vars[
  vars$name %in% c("B01001_001E", "B19013_001E"),
  c("name", "label", "concept")
]
```

```{r geographies}
tc_geography("acs/acs5", 2024)[1:10, c("geography", "summary_level")]
```

```{r tables}
tc_tables("acs/acs5", 2024)[1:5, c("name", "description")]
```

```{r search}
tc_search_variables("acs/acs5", 2024, query = "median household income")[
  1:5,
  c("name", "label", "concept")
]
```

## Migration flows

```{r flows}
tc_get_flows(
  geography = "county",
  year = 2018,
  state = "NY",
  county = "001",
  geometry = "destination"
)
```

## Other products

```{r pep}
tc_get_pep(
  year = 2021,
  dataset = "population",
  variables = "POP_2021",
  geography = "state",
  state = c("NY", "Delaware")
)
```

```{r cbp}
tc_get_cbp(
  year = 2021,
  variables = "ESTAB",
  geography = "state",
  state = c("NY", "DE")
)
```

## Time-series datasets

The package also supports discovery-catalog time-series endpoints:

```{r timeseries}
tc_get_timeseries(
  dataset = "intltrade/exports/hs",
  variables = "ALL_VAL_MO",
  time = "2024-01",
  predicates = list(CTY_CODE = "2010")
)
```

## Optional geometry with tinytiger

When `geometry = TRUE`, `tinycensus` fetches the tabular result first and then joins matching geometry from `tinytiger`.
Set `keep_geo_vars = TRUE` if you want the original geometry attributes as well.

```{r geometry}
tc_get_acs(
  year = 2024,
  variables = "B01001_001E",
  geography = "state",
  state = c("NY", "Delaware"),
  geometry = TRUE,
  keep_geo_vars = TRUE
)
```

## Current scope

`tinycensus` is currently focused on aggregate Census API datasets.
That means the package is aimed at products like ACS, decennial, PEP, CBP, and similar tabular endpoints available through the Census discovery feed.
Microdata-specific ergonomics are not the focus of the current development version.
