---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tinycensus

<!-- badges: start -->
<!-- badges: end -->

`tinycensus` is a lightweight, metadata-driven interface to the [US Census Bureau API](https://www.census.gov/data/developers.html).
It is designed to stay small while still covering a broad set of aggregate datasets, including ACS, decennial census products, population estimates, county business patterns, time-series endpoints, and other datasets exposed through the Census discovery catalog.

The package is built around a generic query engine plus a small set of convenience wrappers.
The goal is to make common Census workflows easy without pulling in a large dependency stack.

## Installation

You can install the development version of tinycensus from GitHub with:

``` r
# install.packages("pak")
pak::pak("christopherkenny/tinycensus")
```

## What it does

- Discovers Census datasets from the live API catalog
- Retrieves metadata for variables, groups, geographies, and example queries
- Pulls tabular results with one generic interface
- Provides wrappers for common products like ACS and decennial census
- Accepts flexible geography inputs like `"NY"`, `"New York"`, and `"36"`
- Returns optional `sf` output through `tinytiger`

## API key

You can use many endpoints without a key for light exploration, but a Census API key is recommended for regular usage.

``` r
tinycensus::tc_set_key("YOUR-KEY")
```

To write the key to a `.Renviron` file for future sessions, set `install = TRUE`
and supply the target path explicitly:

``` r
tinycensus::tc_set_key(
  "YOUR-KEY",
  install = TRUE,
  r_env = file.path(Sys.getenv("HOME"), ".Renviron")
)
```

You can also check whether a key is available:

```{r key}
library(tinycensus)

tc_has_key()
```

## Discover datasets

```{r datasets}
tc_datasets(year = 2024, family = "acs")[1:5, c("year", "dataset", "title")]
```

## Flexible geography inputs

State inputs can be abbreviations, names, or FIPS codes directly in the query interface:

```{r normalize}
tc_get_acs(
  year = 2024,
  variables = "B01001_001E",
  geography = "state",
  state = c("NY", "Delaware", "36")
)
```

County normalization is geography-vintage aware.
That means `tinycensus` uses the county definitions implied by the dataset rather than assuming that the request year alone determines valid county codes.
For example, ACS 2024 uses current county equivalents, while decennial 2020 products keep 2020 county definitions.
If you need to override that behavior for a specific dataset, use `geography_vintage =`.

## Pull data with wrappers

The ACS wrapper makes common requests compact:

```{r acs}
tc_get_acs(
  year = 2024,
  variables = c("B01001_001E", "B19013_001E"),
  geography = "state",
  state = c("NY", "Delaware")
)
```

You can do the same with a decennial dataset:

```{r decennial}
tc_get_decennial(
  year = 2020,
  dataset = "pl",
  variables = "P1_001N",
  geography = "county",
  state = "Delaware"
)
```

## Use the generic interface

`tc_get()` is the core engine. Wrappers like `tc_get_acs()` mainly prefill dataset details.

```{r generic}
tc_get(
  dataset = "acs/acs5",
  year = 2024,
  variables = "B01001_001E",
  geography = "state",
  state = c("NY", "Delaware")
)
```

## Inspect metadata

Metadata helpers make it easier to explore unfamiliar datasets before you query them.

```{r metadata}
vars <- tc_variables("acs/acs5", 2024)
vars[
  vars$name %in% c("B01001_001E", "B19013_001E"),
  c("name", "label", "concept")
]
```

```{r geographies}
tc_geography("acs/acs5", 2024)[1:10, c("geography", "summary_level")]
```

## Time-series datasets

The package also supports discovery-catalog time-series endpoints:

```{r timeseries}
tc_get_timeseries(
  dataset = "intltrade/exports/hs",
  year = NULL,
  variables = "ALL_VAL_MO",
  predicates = list(time = "2024-01", CTY_CODE = "2010")
)
```

## Optional geometry with tinytiger

When `geometry = TRUE`, `tinycensus` fetches the tabular result first and then joins matching geometry from `tinytiger`.

```{r geometry}
tc_get_acs(
  year = 2024,
  variables = "B01001_001E",
  geography = "state",
  state = c("NY", "Delaware"),
  geometry = TRUE
)
```

## Current scope

`tinycensus` is currently focused on aggregate Census API datasets.
That means the package is aimed at products like ACS, decennial, PEP, CBP, and similar tabular endpoints available through the Census discovery feed.
Microdata-specific ergonomics are not the focus of the current development version.
